---
title: "PSTAT100 Data Science Concepts and Analysis"
subtitle: "\\Large Lab 2: Tidy Data and Data Transformations"
author: "Yuting Ma"
format:
  html:
    documentclass: scrartcl
    pdf-engine: xelatex
    fontsize: 11pt
    geometry: margin=0.7in
    highlight-style: monochrome
number-sections: false
execute:
  eval: false
  echo: true
  warning: false
  message: false
mainfont: Latin Modern Roman
monofont: Latin Modern Mono
---

**Name:** __________________________  **Section:** __________________

# Objectives

This lab covers the following topics:

- **Tidy datasets:** identifying variables and observational units.
- **Reshaping tables:** `pivot_longer()` and `pivot_wider()`.
- **Selecting data:** `filter()`, `select()`, and `slice()`.
- **Grouped summaries:** `group_by()` and `summarize()`.

**Student worksheet**



# 1. Tidy Datasets 

\newpage

## 1.1 Creating the Example Dataset

These are fictional scores out of 10. Each student belongs to section A or B;
`NA` means the score was not recorded. Our question is about scores by quiz,
so we choose **one student taking one quiz** as the observational unit.

```{r}
library(tidyverse)
scores_wide <- tibble(
  student = c("Ana", "Ben", "Cam", "Dee"),
  section = c("A", "A", "B", "B"),
  quiz1 = c(8, 6, 9, 7),
  quiz2 = c(10, NA, 8, 9)
)
```
| student | section | quiz1 | quiz2 |
|:--|:--|--:|--:|
| Ana | A | 8 | 10 |
| Ben | A | 6 | NA |
| Cam | B | 9 | 8 |
| Dee | B | 7 | 9 |

For the student-quiz unit, `quiz1` and `quiz2` are values of a quiz variable
stored as column names. We want columns `student`, `section`, `quiz`, `score`.
If the unit were a student with two distinct score variables, a wide layout
could also be appropriate.

## Hands-on Exercise 1: Recognize the unit

A table has columns `species`, `body_weight_kg`, and `brain_weight_g`, with
one row per species. Is it tidy for studying brain and body weight by species?
State the observational unit and explain whether you need to pivot.

\vspace{0.45in}



\newpage

# 2. Reshaping Tables 

`pivot_longer()` moves column names into a variable. `pivot_wider()` moves
values from one variable into column names. These are **tidyr** functions.

## 2.1 Pivoting Longer

```{r}
scores_long <- scores_wide |>
  pivot_longer(
    cols = starts_with("quiz"),
    names_to = "quiz",
    values_to = "score"
  )
scores_long
```
Result: **8 rows, 4 columns**, in student order with quiz1 then quiz2.
The first four rows are:

| student | section | quiz | score |
|:--|:--|:--|--:|
| Ana | A | quiz1 | 8 |
| Ana | A | quiz2 | 10 |
| Ben | A | quiz1 | 6 |
| Ben | A | quiz2 | NA |

`cols` chooses columns to reshape. `names_to` and `values_to` are quoted
because they provide **new column names**. The missing score is retained.

## 2.2 Pivoting Wider

```{r}
scores_long |>
  pivot_wider(
    id_cols = c(student, section),
    names_from = quiz,
    values_from = score
  )
```
Result: the original **4-row, 4-column** table in Section 1. Here `quiz` and
`score` are existing columns, so they can be unquoted. Each student-quiz
combination occurs once; duplicate combinations would need attention.

## Hands-on Exercise 2: Both directions

**a.** Rewrite the longer pivot using `cols = c(...)` instead of a helper.
Use new column names `assessment` and `points`; save as `scores_alt`.

\textcolor{red}{\textbf{Demo Codes:}}

::: callout

```{r}
scores_alt <- scores_wide |>
  pivot_longer(cols = c(____, ____),
               names_to = "____", values_to = "____")
```

:::

**b.** Convert `scores_alt` back to wide. How many rows will the result have?

\textcolor{red}{\textbf{Demo Codes:}}

::: callout

```{r}
scores_alt |>
  pivot_wider(id_cols = c(student, section),
              names_from = ____, values_from = ____)
```

:::


Rows: __________

\newpage

# 3. Selecting Rows and Columns 

These **dplyr** functions choose rows or columns.

| Function | Chooses | Example |
|:--|:--|:--|
| `filter()` | Rows meeting a condition | `filter(score >= 9)` |
| `select()` | Columns by name or helper | `select(student, score)` |
| `slice()` | Rows by position in current order | `slice(1:2)` |

Use unquoted column names and quoted text values. `==` compares values;
commas in `filter()` mean AND; `|` means OR.

## 3.1 Filtering and Selecting

```{r}
scores_long |>
  filter(section == "A", score >= 9) |>
  select(student, quiz, score)
```
| student | quiz | score |
|:--|:--|--:|
| Ana | quiz2 | 10 |

`filter()` keeps only TRUE conditions; missing comparisons are dropped.
Use `is.na(score)` to find missing scores.

## 3.2 Slicing by Position

```{r}
scores_wide |>
  select(student, starts_with("quiz")) |>
  slice(c(1, 3))
```
**Result:** Ana (quiz1 = 8, quiz2 = 10) and Cam (quiz1 = 9, quiz2 = 8).

`slice()` uses current row order, not score size; grouped slicing works
within each group.

## Hands-on Exercise 3: Write code and predict the result

**a.** From `scores_long`, keep section B scores of at least 9; return only
`student` and `quiz`. Write the code and the student-quiz pairs returned.

\textcolor{red}{\textbf{Demo Codes:}}

::: callout

```{r}
scores_long |>
  filter(____________________________) |>
  select(____________________________)
```

:::


Returned pairs: _______________________________________________

\vspace{0.15in}
**b.** From `scores_wide`, keep only columns `student` and `quiz2`, then
keep row positions 2 and 4. Write code. Which two students remain?

\textcolor{red}{\textbf{Demo Codes:}}

::: callout

```{r}
# Your code:
_______________________________________________________________
```

:::


Students: __________________________

\newpage

# 4. Grouping and Summarizing 

`group_by()` tells later operations which rows belong together; it does not
itself reduce the number of rows. `summarize()` with one summary value per
group returns one row per group. `summarise()` is an equivalent spelling.

## 4.1 Summarizing by Section

```{r}
scores_long |>
  group_by(section) |>
  summarize(
    n_rows = n(),
    n_recorded = sum(!is.na(score)),
    mean_score = mean(score, na.rm = TRUE),
    .groups = "drop"
  )
```
| section | n_rows | n_recorded | mean_score |
|:--|--:|--:|--:|
| A | 4 | 3 | 8.00 |
| B | 4 | 4 | 8.25 |

`n()` counts **rows**, including missing scores. Section A's mean is
$(8+10+6)/3=8$, not divided by 4. `na.rm = TRUE` excludes missing scores
from the calculation, not from the dataset. Without it, A's mean is `NA`.
`.groups = "drop"` makes the result ungrouped.

**Think about the denominator:** Here rows are student-quiz observations,
not students. Without `group_by()`, `summarize()` would give one overall row.

## Hands-on Exercise 4: Two grouping variables

Compute the **mean recorded score** separately for each section and quiz.
Also report the number of recorded scores. Use `scores_long`.

\textcolor{red}{\textbf{Demo Codes:}}

::: callout

```{r}
scores_long |>
  group_by(__________________________) |>
  summarize(
    n_recorded = _____________________,
    mean_score = _____________________,
    .groups = "drop"
  )
```

:::

Fill in the expected output. Then explain why A-quiz2 has only one
recorded score even though section A has two students.

| section | quiz | n_recorded | mean_score |
|:--|:--|:--|:--|
| A | quiz1 | __________ | __________ |
| A | quiz2 | __________ | __________ |
| B | quiz1 | __________ | __________ |
| B | quiz2 | __________ | __________ |

Explanation: __________________________________________________

______________________________________________________________

\newpage

# 5. Independent Practice 

Work alone or with a partner. Use the setup on Section 1 and `scores_long` from
Section 2. Write code and predict results before checking in R.

## Hands-on Exercise 5: Build a complete pipeline

Starting with `scores_wide`, (1) pivot to `quiz` and `score`; (2) keep only
quiz2 rows with recorded scores; (3) calculate the mean score and number of
recorded scores by section; (4) keep only sections with a mean of at least 9.
Call the count column `n_recorded` and the mean column `mean_score`.

\textcolor{red}{\textbf{Demo Codes:}}

::: callout

```{r}
# Your pipeline (use the space below):
```

:::


\vspace{1.0in}

Which section(s) remain? Give the count and mean. Why is the last filter
after `summarize()` rather than before it?

______________________________________________________________

______________________________________________________________

\vspace{0.2in}
## Hands-on Exercise 6: Diagnose and repair 

A classmate wants the mean score by section but writes:

\textcolor{red}{\textbf{Demo Codes:}}

::: callout

```{r}
scores_long |> summarize(mean_score = mean(score))
```

:::

Identify **two problems**, write corrected code, and predict its output.

\textcolor{red}{\textbf{Demo Codes:}}

::: callout

```{r}
# Problems: ___________________________________________________
# Corrected code:
```

:::


\vspace{0.45in}


Output: _______________________________________________________


______________________________________________________________

