---
title: "PSTAT100 Data Science Concepts and Analysis"
subtitle: "\\Large Lab 1: Dataframe Overview with R"
author: "Claire Gan"
date: 1-13-2026
date-format: full
format: 
  pdf:
    documentclass: scrartcl
    fontsize: 11pt
number-sections: false
execute: 
  eval: true
  echo: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message =  FALSE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(error =  FALSE)
bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}

# Install necessary libraries if they aren't installed
if (!require(dplyr)) install.packages("dplyr")
if (!require(tidyr)) install.packages("tidyr")
if (!require(ggplot2)) install.packages("ggplot2")
```



R is a programming language widely used for statistical computing and data analysis. In this tutorial, we will focus on the `dplyr` package for data manipulation.

# Objectives

This lab covers the following topics:

- **Dataframe basics**
  - Creating dataframes
  - Dataframe indexing and attributes
  - Adding, removing, and renaming variables
- **Operations on dataframes**
  - Slicing (selecting rows and columns)
  - Filtering (selecting rows that meet certain conditions)


```{r}
# Load libraries
library(dplyr)
library(ggplot2)
```

# 1. Creating DataFrames & Basic Manipulations

A **dataframe** in R is a table where each column has a specific type (numeric, character, etc.). It has two main indices:

-   **Column index**: Labels for each column (usually strings).

-   **Row index**: Ordinal numbers representing each row.

In this tutorial, we will cover:

1.  Creating dataframes from scratch.

2.  Retrieving dataframe attributes.

3.  Indexing dataframes.

4.  Adding, removing, and renaming columns.

We will use the example dataset: **`fruit_info`**, which contains information about different fruits.

## 1.1 Creating DataFrames from Scratch

In R, we use the `data.frame()` function to create a dataframe. Let’s create a dataframe for the **`fruit_info`** dataset.

```{r}
# Creating a dataframe using a list
fruit_info <- data.frame(
  Fruit = c("Apple", "Banana", "Cherry", "Date", "Elderberry"),
  Color = c("Red", "Yellow", "Red", "Brown", "Purple"),
  Weight = c(150, 120, 10, 25, 5)
)

# Display the dataframe
print(fruit_info)
```


## 1.2 Retrieving Attributes

Once the dataframe is created, you may want to check its attributes like the column names, the number of rows, and the number of columns.

```{r}
# Checking the number of rows and columns
nrow(fruit_info)
ncol(fruit_info)

# Displaying column names
colnames(fruit_info)

# Displaying the structure of the dataframe
str(fruit_info)
```

## 1.3 Indexing DataFrames

In R, you can access the data in a dataframe in various ways.

### [Column Indexing]{.underline}

You can access columns by name or by column index.

```{r}
# Accessing columns by name index
fruit_info$Fruit

# Accessing columns by number index
fruit_info[,1]
```

### [Row Indexing]{.underline}

You can access rows by number or by condition.

```{r}
# Accessing rows by number
fruit_info[1, ]
fruit_info[1:3, ]
fruit_info[c(2,4,6), ]

# Accessing rows by condition (Weight > 100)
fruit_info[fruit_info$Weight > 100, ]
```


## 1.4 Adding, Removing, and Renaming Columns

### [Adding a Column]{.underline}

You can add a new column by assigning values to a new column name.

```{r}
# Adding a new column for price (just an example)
fruit_info$Price <- c(1.2, 0.5, 2.0, 3.5, 4.0)

# View updated dataframe
print(fruit_info)
```

### [Removing a Column]{.underline}

You can remove a column using `select()` from the `dplyr` package.

```{r}
# Removing the Score column
library(dplyr)
fruit_info <- select(fruit_info, -Price)

# View updated dataframe
print(fruit_info)
```


### [Renaming Columns]{.underline}

Use `rename()` to change column names.

```{r}
# Renaming columns
fruit_info <- rename(fruit_info, Fruit_Name = Fruit)

# View updated dataframe
print(fruit_info)
```

## Hand-on Exercise 1

### (1) Adding a new column:

Using direct specification, add to the `fruit_info` table a new column called `rank1` containing integers 1, 2, 3, 4, and 5, which express your personal preference about the taste ordering for each fruit (1 is the tastiest; 5 is the least tasty). Make sure that the numbers utilized are unique - no ties are allowed.

`r bfcolor("Demo Codes:", "red")`

::: callout
```{r}
# Adding a new column 'rank1' to express personal preference


# View the updated dataframe

```
:::

### (2) Creating a modified dataframe:

Now, create a new dataframe `fruit_info_mod1` with the same information as `fruit_info`, but with an additional column called `rank2`. We'll start by making a copy of the `fruit_info` dataframe to create the new dataframe `fruit_info_mod1`.

`r bfcolor("Demo Codes:", "red")`

::: callout
```{r}
# Creating a new dataframe 'fruit_info_mod1' as a copy of 'fruit_info'

```
:::

Then, using **name indexing**, add a column called `rank2` to the `fruit_info_mod1` dataframe that contains the same values in the same order as the `rank1` column.

`r bfcolor("Demo Codes:", "red")`

::: callout
```{r}
# Adding 'rank2' to 'fruit_info_mod1' using the same values as 'rank1'

# View the updated dataframe

```
:::

When using the name indexing approach, the array assigned to the new variable must be **the same length** as the number of rows in the dataframe. What if we only assign values to certain rows?

### (3) New column assignment with missing data:

For example, let’s try adding a new column `rank3` and assign values only to the first two rows. 

`r bfcolor("Demo Codes:", "red")`

::: callout
```{r}
# Assign 'rank3' to only the first two rows

# View the updated dataframe with missing values in 'rank3'

```
:::

::: callout
[**Hints:**]{.underline}

-   **Partial Assignment**: We use row-based indexing (`1:2`) to assign values only to the first two rows of `rank3`. The remaining rows in `rank3` will have `NA` values by default because we didn't assign values to them.

-   **Missing Values**: The unassigned rows will contain missing values (`NA`).
:::


### (4) Checking missing entry:

We can detect missing values using the `is.na()` method in R, which will return a logical dataframe indicating whether an entry is missing.

`r bfcolor("Demo Codes:", "red")`

::: callout
```{r}
# Detect missing values in the dataframe

# Print missing data frame

```
:::

::: callout
[**Hints:**]{.underline}

-   **Missing Data Check**: The `is.na()` function checks for missing (`NA`) values in the dataframe. It returns a logical dataframe where `TRUE` indicates missing values.

-   **Visualization**: This is helpful for identifying rows or columns with missing data.
:::

It’s often more helpful to see if any column has missing values. You can do this by appending `any()` to the `is.na()` function, which will return `TRUE` if there are any missing values in the column.

`r bfcolor("Demo Codes:", "red")`

::: callout
```{r}
# Check if any column contains missing values


# Check for missing data in each column


# Print the result

```
:::

::: callout
[**Hints:**]{.underline}

-   **Missing Data Detection**: We use `colSums(is.na(fruit_info_mod1))` to check for missing data in each column. This will give us the number of missing entries per column.
-   **Helpful for Clean-Up**: This helps identify which columns may need further attention or data cleaning.
:::

### (5) Removing column(s) with missing entry:

Once we've finished with some columns, we might want to remove them. In R, we can use `select()` from the `dplyr` package or direct indexing to remove columns.

`r bfcolor("Demo Codes:", "red")`

::: callout
```{r}
# Removing a column using dplyr's select()
library(dplyr)


# View the updated dataframe

```
:::

::: callout
[**Hints:**]{.underline}

-   **Removing Columns**: We used the `select()` function from `dplyr` to remove the `rank3` column from the dataframe. The negative sign (`-`) indicates that the column should be removed.

-   **Data Cleaning**: This is a common operation when you no longer need certain columns in the dataset.
:::


# 2. Operations on Data Frames

With some basics in place, here you'll see how to perform subsetting operations on data frames that are useful for tidying up datasets.

- **Slicing**: Selecting columns or rows in chunks or by position.  
- **Filtering**: Selecting rows that meet certain criteria.  

We will illustrate these operations using a dataset comprising counts of baby names born in California each year from 1990 to 2018.

```{r}
# Import baby names data
# (Put the csv file in the same folder as this .qmd file)
baby_names <- read.csv("baby_names.csv")

# Preview first few rows
head(baby_names)
```

::: callout
[**Important note:**]{.underline}

If you received some Error messages saying *Error in file(file, "rt") : cannot open the connection*, 
this usually means R cannot find your data file.  
Make sure to put the `baby_names.csv` file in the **same folder** as this `.qmd` file.  
:::

## Hand-on Exercise 2

### (1) Checking dimensions:

Check the dimensions of the `baby_names` dataset and store them in `dimensions_baby_names`.
In R, we use the `dim()` function to get the dimensions of a dataframe.

`r bfcolor("Demo Codes:", "red")`

::: callout
```{r}
# Get the dimensions of the 'baby_names' dataframe

# Print the dimensions

```
:::

### (2) Counting distinct years:

Count the number of occurrences of each distinct year in `baby_names`.
Store the counts as `occur_per_year` (ordered by year), and the total number of years as `num_years`.

`r bfcolor("Demo Codes:", "red")`

::: callout
```{r}
# Count occurrences of each year

# Sort the occurrences by year 

# Get the number of years represented in the dataset

```
:::

::: callout
[**Notes:**]{.underline}

-   **Sorting by Year**: We first count the occurrences of each year with the `table()` function, which gives us a named vector. The `order()` function is then used to sort the vector by the year, which is achieved by converting the names of the `occur_per_year` (which are stored as character strings) into numeric values with `as.numeric()`.

-   **Number of Years**: The `unique()` function is used to get the distinct years, and `length()` gives the number of unique years.
:::


## 2.1 Slicing: Selecting Rows and Columns

In this section, we'll cover two primary ways to slice a dataframe:

-   **Using index names** to specify rows and columns.

-   **Using integer positions** to specify rows and columns.

These methods are very useful for subsetting data and inspecting different portions of a dataframe.

### [1. Slicing with Index Names]{.underline}

To slice a dataframe by index names (column and row names), you can use the `[]` operator in R.
You can select a specific entry by specifying both the row and column names.

```{r}
# Selecting a single row and column by index name
baby_names[2, 'Name']
```

-   **Single Index**: In this example, we select the second row and the `Name` column from the `baby_names` dataframe using row index `2` and column name `'Name'`.

### [2. Slicing with a List of Indices]{.underline}

You can also select multiple rows and specific columns by passing a **list** of indices or names.

```{r}
# Selecting multiple rows and specific columns by index names
baby_names[c(2, 3), c('Name', 'Count')]
```

-   **List of Indices**: We use `c(2, 3)` to select rows 2 and 3, and `c('Name', 'Count')` to select the columns `'Name'` and `'Count'`.

### [3. Slicing with Consecutive Indices]{.underline}

You can select a range of rows or columns using the colon (`:`) operator.

```{r}
# Selecting rows 2 to 10 and columns from 'Year' to 'Count'
select(baby_names[2:10,], Year:Count)
```

-   **Consecutive Indices**: The `2:10` syntax selects rows 2 through 10, and `'Year':'Count'` selects all columns from `Year` to `Count`.

-   **`select(Year:Count)`**: This approach uses `dplyr`'s `select()` function to select columns starting from `'Year'` and ending at `'Count'`. It properly handles the selection of consecutive columns by name, ensuring that no errors occur.

### [4. Slicing with Integer Positions]{.underline}

In R, you can also slice data by integer positions using the `[,]` operator.
To select a specific entry by position, you can specify row and column indices.

```{r}
# Selecting a specific entry by position (row 2, column 3)
baby_names[2, 3]
```

-   **Single Position**: The `baby_names[2, 3]` selects the entry in the second row and third column by position.

### [5. Slicing with a List of Positions]{.underline}

You can select multiple rows and columns by specifying their integer positions.

```{r}
# Selecting multiple rows and columns by position
baby_names[c(2, 3), c(3, 4)]
```

-   **List of Positions**: The `c(2, 3)` selects the second and third rows, and `c(3, 4)` selects the third and fourth columns by position.

### [6. Slicing with Consecutive Positions]{.underline}

You can also slice consecutive rows or columns by specifying a range of positions.

```{r}
# Selecting rows 2 through 11 and columns 2 through 5
baby_names[2:11, 2:5]
```

-   **Consecutive Positions**: The `2:11` selects rows 2 through 11, and `2:5` selects columns 2 through 5 based on their integer positions.


## 2.2 Filtering

Filtering is the process of sifting out rows according to a criterion. In R, this can be accomplished using logical indexing, which creates a vector of `TRUE`s and `FALSE`s based on a comparison. Let's walk through a simple example:

### Filter Names Based on Occurrence Count

Let’s say we want to filter out all names with fewer than 1000 occurrences. First, we can define a logical vector that checks the `Count` column.

```{r}
# Create a logical vector where TRUE means Count is greater than 1000
arr <- baby_names$Count > 1000

# View the logical array
head(arr, sum(arr)*0.05)
```

Once we have the logical array, we can apply it to filter the dataframe and select only the rows where `Count` is greater than 1000.

```{r}
# Filter the dataframe using the logical array
baby_names_filtered <- baby_names[arr, ]

# View the filtered dataframe
head(baby_names_filtered)
```

**Checking Dimensions Before and After Filtering**

To get a sense of how many rows were filtered, we can compare the dimensions of the original and filtered dataframes.

```{r}
# Compare dimensions before and after filtering
print(dim(baby_names))
print(dim(baby_names_filtered))
```

Notice that the filtered dataframe is much smaller than the overall dateframe -- only about **2000** rows correspond to a name occurring more than 1000 times in a year for a gender.

### [Common Comparison Operators]{.underline}

Here are some commonly used comparison operators in R for filtering data:

| Symbol | Usage    | Meaning                              |
|--------|----------|--------------------------------------|
| `==`   | `a == b` | Does `a` equal `b`?                  |
| `<=`   | `a <= b` | Is `a` less than or equal to `b`?    |
| `>=`   | `a >= b` | Is `a` greater than or equal to `b`? |
| `<`    | `a < b`  | Is `a` less than `b`?                |
| `>`    | `a > b`  | Is `a` greater than `b`?             |
| `!`    | `!p`     | Negates `p` (logical NOT)            |
| `&`    | `p & q`  | `p` AND `q`                          |
| `|`    | `p | q`  | `p` OR `q`                           |

What if instead you wanted to filter using multiple conditions? Here's an example of retrieving rows with counts exceeding 1000 for only the year 2001:

### Filtering with Multiple Conditions

You can combine multiple conditions to filter the dataframe in more complex ways. For example, let's retrieve all names where `Count` exceeds 1000 and the `Year` is 2001:

```{r}
# Filter using two conditions
filtered_data <- baby_names[(baby_names$Year == 2001) & (baby_names$Count > 1000), ]

# View the filtered data
head(filtered_data, dim(filtered_data)[1]*0.1)
```

-   **Combining Conditions:** In this example, we combine two conditions using the `&` operator. We filter for rows where `Year` equals 2001 and `Count` is greater than 1000.

## Hand-on Exercise 3:

In R, you can also slice rows based on specific conditions by using logical indexing. For example, you can select rows where a certain column satisfies a condition.

### (1.) Selecting Rows with `Count` Greater than 100

Slice rows where 'Count' is greater than 100 from the baby_name dataframe

`r bfcolor("Demo Codes:", "red")`

::: callout

```{r}
# Slice rows where 'Count' is greater than 100

```
:::

### (2.) Slicing Rows Based on Column Value Range

Selecting Rows for Years 2000 to 2005 from the baby_name dataframe.

`r bfcolor("Demo Codes:", "red")`

::: callout

```{r}
# Slice rows where the 'Year' is between 2000 and 2005

```
:::
-   **Range Condition**: Here, we use a logical condition to select rows where the `Year` is between 2000 and 2005 (inclusive). The logical condition is used within the indexing brackets to filter rows.


